---
title: "tidyweek1test"
author: "BK"
date: "03/26/2018"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    number_section:  true
    theme: cerulean
    highlight:  tango
       
editor_options: 
  chunk_output_type: inline
---

JK: I tend to keep all of my setup/library in a single chunk, and inline custom functions in a single chunk. Focus on using the auto-formatting code options in Rstudio.  Helps keep things in a nice consistent style, and makes things easy to read. I would REALLY recommend using the `here` package, it makes a lot of pathway stuff in Rmarkdown easier.
```{r functions, include = FALSE, echo = FALSE}

is_installed <- function(pkg) is.element(pkg, installed.packages()[,1])

loadLibrary <- function(pkg) {
 if(!is_installed(pkg)) {
   install.packages(pkg)
 } 
library(pkg,character.only = TRUE, quietly = TRUE, verbose = FALSE)
}
```

JK: I like the functional approach.  I made your loadLibrary for loop into a function that you can use with apply or purrr functions.  This follows the standard R paradigm of applying functions to lists or vectors.
```{r setup, include=FALSE}

# Set global chunk options
knitr::opts_chunk$set(echo = TRUE)

packages <- c("tidyverse","knitr","DT","kableExtra","janitor", "lubridate", "data.table", "here", "purrr")

invisible(lapply(packages, loadLibrary))

```

JK: the rm(list = ls()) isn't the best way to clean your workspace.  https://www.tidyverse.org/articles/2017/12/workflow-vs-script/
## **Preparing for the analysis**
### Cleaning work space


## Project goals
This project requires the learn and use the skills to use [tidyverse packages](https://www.tidyverse.org/packages/):  

  1. Import data using [readr](http://readr.tidyverse.org/) package from csv file   
  2. Use [tidyr](http://tidyr.tidyverse.org/) package and [dplyr](http://dplyr.tidyverse.org/) package to   tidy the data  
  3. Perform Exploratory Data Analysis (EDA) to explore a specific research question to explore the relationship between mental health and mass shooting across years


## Importing Data3
The data is from [World Bank](https://datacatalog.worldbank.org/dataset/sustainable-development-goals). TRelevant indicators drawn from the World Development Indicators, reorganized according to the goals and targets of the Sustainable Development Goals (SDGs). These indicators may help to monitor SDGs, but they are not always the official indicators for SDG monitoring.

## Data manipulation and cleaning

### Subsetting 

JK: 
   * Make use of chunk labels!  They help immensely in navigating. 
   * You load janitor! make use of `clean_names` and `remove_empty`.  These are amazing helpers!
   * If speed isn't a big issue, `fread` vs `read_csv` isn't worth loading `data.table`
```{r load_data, message=FALSE}

path_data <- here("data/SDG.rds")

# SDG <- read_csv(path_data) %>% 
#   clean_names() %>% 
#   remove_empty(c("cols", "rows"))
# 
# saveRDS(SDG, here('data/SDG.rds'))

SDG <- readRDS(path_data)

dim(SDG)
#remove X in the column names such as X1960, X1961...
names(SDG) <- sub("^x", "", names(SDG))
names(SDG)
    
```

Mortality rate, under-5 (per 1,000 live births)"="SH.DYN.MORT"
"GDP growth (annual %)"="NY.GDP.MKTP.KD.ZG"
"GNI growth (annual %)"="NY.GNP.MKTP.KD.ZG"
"Number of people spending more than 10% of household consumption or income on out-of-pocket health care expenditure"="SH.UHC.OOPC.10.TO"
"Number of people spending more than 25% of household consumption or income on out-of-pocket health care expenditure"="SH.UHC.OOPC.25.TO"
"Nurses and midwives (per 1,000 people)"="SH.MED.NUMW.P3"
"Women making their own informed decisions regarding sexual relations, contraceptive use and reproductive health care  (% of women age 15-49)"=	"SG.DMK.SRCR.FN.ZS"
"Methodology assessment of statistical capacity (scale 0 - 100)"="IQ.SCI.MTHD"
"Periodicity and timeliness assessment of statistical capacity (scale 0 - 100)"="IQ.SCI.PRDC"
"Source data assessment of statistical capacity (scale 0 - 100)"="IQ.SCI.SRCE"

JK: Running the janitor cleaning steps will eliminate the select and rename steps.  This lets you filter to what you want. Good using `%in%`!  Remember your chunk label!
```{r}

  df <- SDG %>%
    #     select ("Indicator Name","Country Name","Country Code","Indicator Code",
    #             c(`1960`:`2017`)) %>%
    # rename (indicator="Indicator Name",
    #         country="Country Name",
    #         code="Country Code",
    #         indicator_code="Indicator Code") %>%
  filter ( indicator_code %in%   c("IQ.SCI.MTHD",
                                   "IQ.SCI.PRDC",
                                   "IQ.SCI.SRCE",
                                   "NY.GDP.MKTP.KD.ZG","SH.DYN.MORT",
                                   "NY.GNP.MKTP.KD.ZG","SH.MED.NUMW.P3",
                                   "SH.UHC.OOPC.25.TO","SH.UHC.OOPC.10.TO"))

kable (head(df,2))
dim(df)


```


### What is a tidy data set?
[Hadley Wickham's Tidy Data](http://vita.had.co.nz/papers/tidy-data.pdf) framework  suggests tidy data must have three properties:  

   + Each variable forms a column.    
   + Each observation forms a row.    
   + Each type of observational unit forms a table  

**What is the observational unit in this mass_shooting data frame?**

*Is the observational country? or other variables  ?*
In this example, the data is alreay in not in tidy format. 
## Using tidyr functions

JK:  While this method of filtering works, and this is fantastic way of undetsanding the steps you need to filter one subset of the data, you are repeating yourself.  One of the fundamental rules of programming is DRY (don't repeat yourself).  If you find yourself cutting and pasting the step of code, make it into a function and call it on a list or nested frame.  The benefit is cleaner and clearer code for others to follow you analysis.  How would you take this and make it into a function and apply it to the data?  Or could you do this all at once?  Let me know if you need some hints!

You also have some variable name inconsistencies in the code!
```{r}
#method variable tidying

df1<-df %>% 
  #filtering on "Methodology assessment of statistical capacity (scale 0 - 100)"="IQ.SCI.MTHD" 
  filter ( indicator_code=="IQ.SCI.MTHD") %>%
   select (-c(indicator, code)) %>%
     gather ("year", "periodicity" , c(`1990`:`2017`) )%>%
        select (-indicator_code)%>%
        arrange (country, year)
     

df2<-df %>% 
 #filtering on Periodicity and timeliness assessment of statistical capacity (scale 0 - 100)"="IQ.SCI.PRDC"
  filter ( indicator_code=="IQ.SCI.PRDC") %>%
  select (-c(indicator, code)) %>%
     gather ("year", "periodicity" , c(`1990`:`2017`) )%>%
        select (-indicator_code)%>%
        arrange (country, year)


df3<-df %>% 
#filtering on Methodology assessment of statistical capacity (scale 0 - 100)"="IQ.SCI.SRCE"
  filter ( indicator_code=="IQ.SCI.PRDC") %>%
  select (-c(indicator, code)) %>%
     gather ("year", "assessment" , c(`1990`:`2017`) )%>%
        select (-indicator_code) %>%
        arrange (country, year)

df4<-df %>% 
#filtering on Mortality rate, under-5 (per 1,000 live births)"="SH.DYN.MORT"
  filter ( indicator_code=="SH.DYN.MORT") %>%
  select (-c(indicator, code)) %>%
     gather ("year", "mortality" , c(`1990`:`2017`) )%>%
        select (-indicator_code)%>%
        arrange (country, year)

goal2 <- df1 %>%
full_join (df2,   by=c("year", "country")) %>%
full_join (df3,   by=c("year", "country")) %>%
inner_join(df4,   by=c("year", "country"))
goal2$year <- as.numeric(as.character(goal2$year))

datatable(goal2)


```

## EDA

### Research Question1

RQ1: Do Mortality rate, under-5 (per 1,000 live births) increase or decrease over the years in each country?

```{r  ,fig.height = 15, fig.width = 15, fig.align = "center" , message= FALSE}
ggplot (goal2, aes(x=year, y=mortality, color=country))+ geom_point () +
    labs(x="Year" , y="Mortality rate, under-5 (per 1,000 live births)") + facet_wrap (~country) +
   scale_x_continuous (breaks=seq(1990,2016,15), limits = c(1990,2016))+
   # scale_y_continuous (breaks=seq(0,60,10), limits = c(0 , 60))+
  ggtitle ("figure1.Do Mortality rate, under-5 (per 1,000 live births) increase or decrease over the years in each country?")+
  theme(legend.position = 'none')
```



### Research Question2

RQ2: How does the  source data assessment of statistical capacity change across years in each country?

```{r  ,fig.height = 15, fig.width = 15, fig.align = "center" , message= FALSE}

RQ2<-ggplot (goal2, aes(x=year, y=assessment, color=country))+ geom_jitter () +
    labs(x="Year" , y="Source data assessment of statistical capacity (scale 0 - 100)") + facet_wrap (~country) +
  ggtitle ("figure2.How does the  source data assessment of statistical capacity across years in each country?")+
  theme(legend.position = 'none')
RQ2

```


### Research Question3

RQ3: How does the methodology assessment of statistical capacity (scale 0 - 100) change across years in each country?

```{r  ,fig.height = 15, fig.width = 15, fig.align = "center" , message= FALSE}

RQ3<-ggplot (goal2, aes(x=year, y=method, color=country))+ geom_jitter () +
    labs(x="Year" , y="Methodology assessment of statistical capacity (scale 0 - 100)") + facet_wrap (~country) +
  ggtitle ("figure3.How does the methodology assessment of statistical capacity (scale 0 - 100) change across years in each country?")+
  theme(legend.position = 'none')
RQ3

```

### Research Question4

RQ4: How does the periodicity and timeliness assessment of statistical capacity (scale 0 - 100) change across years in each country?

```{r  ,fig.height = 15, fig.width = 15, fig.align = "center" , message= FALSE}

RQ4<-ggplot (goal2, aes(x=year, y=periodicity, color=country))+ geom_jitter () +
    labs(x="Year" , y="Periodicity and timeliness assessment of statistical capacity (scale 0 - 100)") + facet_wrap (~country) +
  ggtitle ("figure4.How does the periodicity and timeliness assessment of statistical capacity (scale 0 - 100) change across years in each country?")+
  theme(legend.position = 'none')
RQ4

```



##Closing remarks 
RQ1: The Child Mortality in all countries did drop across years from 1990 to 2016 in this data to track the Sustainable Development Goals (SDGs). Angola particulary had a very shorp decrease across years in the number of child mortality under-5 (per 1,000 live births). Similary, Niger, Rwanda, Liberia, Malawi are countries with a sharp decrease in child mortality.

RQ2: Regarding the source data assessment of statistical capacity, most countries have no change in terms of decrease of increase in their capacity across years. 

RQ3: With respect to the methodology assessment of statistical capacity, Serbia made significant sharp increase, similary Seychelles is another country that had made signficant improvement over time.Interestingly, Ethophia, had some non linear relationship on this matter where there was an increase at the beginning then there was a decrease in theri capacity. Overall, most countries were stable in their methodology assessment of statistical capacity improvement.

RQ4: Regarding the periodicity and timeliness assessment of statistical capacity, Serbia, Monenegro, Sierra Leone made significant sharp increase but most countries were stable. 

